Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-05-26 paper/arm.tex
generlize RMW
Added m5a

Even with this small change, the optimal lowering for \armeight{} is unsound
for our semantics.  The optimal lowering maps relaxed access to \LDR/\STR{} and
non-relaxed access to \LDAR/\STLR{} \citep{DBLP:journals/pacmpl/PodkopaevLV19}.
In this section, we consider a suboptimal strategy, which lowers non-relaxed
reads to $(\DMBSY\SEMI\LDAR)$.  Significantly, we retain the optimal lowering
for relaxed access.  In the next section we recover the optimal lowering by
adopting an alternative semantics for \ref{seq-le}.

@@

      \vspace{-.5\baselineskip}
      \begin{proof}
        By way of contradiction, assume $\aEv\xgcb\cEv$.  If $\cEv\xrfx\aEv$
        then by \ref{arm-gcb} we must also have $\cEv\xgcb\aEv$,
        contradicting the assumption that $\rgcb$ is a total order.
        %
        Otherwise that there is some $\bEv\neq\cEv$ such that
        $\bEv\xrfx\aEv$, and therefore $\bEv\xgcb\aEv$.  By transitivity,
        $\bEv\xgcb\cEv$.  By the definition of $\rfr$, we have
        $\aEv\xfr\cEv$.  But this contradicts \ref{arm-local}, since
        $\cEv\xpoloc\aEv$.
        %Applying \ref{arm-gcb-blocks}, we have that either $\cEv\xgcb\bEv$ or $\aEv\xgcb\cEv$.
      \end{proof}
    \end{lemma}
    We show that all the order required in the pomset is also required by
    \armeight{}.  \ref{rf-block} holds since ${\rcb_1}$ is consistent with
    ${\rco_1}$ and ${\rfr_1}$.  As noted \ref{lob-le}, $\rlob$ includes the order
    required by \ref{seq-kappa} and \ref{seq-le-delays}.  We need only show
    that the order removed from \ref{arm-gcb-lob} can also be removed from
    the pomset.  In order for \ref{arm-gcb-lob} to remove order from $\aEv$
    to $\cEv$, we must have $\bEv\xrfx\aEv$ and $\bEv\xpoloc\aEv$ but not
    $\bEv\xlob\cEv$.  Because of our suboptimal lowering, it must be that
    $\aEv$ is a relaxed read; otherwise the $\DMBSY$ would require
    $\bEv\xlob\cEv$.  Thus we know that \ref{seq-le-delays} does not require
    order from $\aEv$ to $\cEv$.  By chaining \ref{read-tau-ind} and
    \ref{write-tau}, any dependence on the read can by satisfied without
    introducing order in \ref{seq-kappa}.
  \end{proof}  
\end{theorem}

@@

  \vspace{-.5\baselineskip}
  \begin{proof}
    We show that all the order required in the pomset is also required by
    \armeight{}.  \ref{rf-block} holds since ${\rcb_1}$ is consistent with
    ${\rco_1}$ and ${\rfr_1}$.  \ref{seq-le-rf-rf} follows from \ref{arm-rfi}.
    As noted \ref{lob-le}, $\rlob$ includes the order required by
    \ref{seq-kappa} and \ref{seq-le-delays-rf}.  
  \end{proof}
\end{theorem}

>>>>

Even with this small change, the optimal lowering for \armeight{} is unsound
for our semantics.  The optimal lowering maps relaxed access to \LDR/\STR{} and
non-relaxed access to \LDAR/\STLR{} \citep{DBLP:journals/pacmpl/PodkopaevLV19}.
In this section, we consider a suboptimal strategy, which lowers non-relaxed
reads to $(\DMBSY\SEMI\LDAR)$.  Significantly, we retain the optimal lowering
for relaxed access.  In the next section we recover the optimal lowering by
adopting an alternative semantics for \ref{seq-le-rf} and \ref{seq-le-delays}.

@@

      \vspace{-.5\baselineskip}
      \begin{proof}
        By way of contradiction, assume $\aEv\xgcb\cEv$.  If $\cEv\xrfx\aEv$
        then by \ref{arm-gcb} we must also have $\cEv\xgcb\aEv$,
        contradicting the assumption that $\rgcb$ is a total order.
        %
        Otherwise that there is some $\bEv\neq\cEv$ such that
        $\bEv\xrfx\aEv$, and therefore $\bEv\xgcb\aEv$.  By transitivity,
        $\bEv\xgcb\cEv$.  By the definition of $\rfr$, we have
        $\aEv\xfr\cEv$.  But this contradicts \ref{arm-local}, since
        $\cEv\xpoloc\aEv$.
        %Applying \ref{arm-gcb-blocks}, we have that either $\cEv\xgcb\bEv$ or $\aEv\xgcb\cEv$.
      \end{proof}
    \end{lemma}
    We show that all the order required in the pomset is also required by
    \armeight{}.  \ref{pom-le-block} holds since ${\rcb_1}$ is consistent with
    ${\rco_1}$ and ${\rfr_1}$.  As noted \ref{lob-le}, $\rlob$ includes the order
    required by \ref{seq-kappa} and \ref{seq-le-delays}.  We need only show
    that the order removed from \ref{arm-gcb-lob} can also be removed from
    the pomset.  In order for \ref{arm-gcb-lob} to remove order from $\aEv$
    to $\cEv$, we must have $\bEv\xrfx\aEv$ and $\bEv\xpoloc\aEv$ but not
    $\bEv\xlob\cEv$.  Because of our suboptimal lowering, it must be that
    $\aEv$ is a relaxed read; otherwise the $\DMBSY$ would require
    $\bEv\xlob\cEv$.  Thus we know that \ref{seq-le-delays} does not require
    order from $\aEv$ to $\cEv$.  By chaining \ref{read-tau-ind} and
    \ref{write-tau}, any dependence on the read can by satisfied without
    introducing order in \ref{seq-kappa}.
  \end{proof}  
\end{theorem}

@@

  \vspace{-.5\baselineskip}
  \begin{proof}
    We show that all the order required in the pomset is also required by
    \armeight{}.  \ref{pom-le-block} holds since ${\rcb_1}$ is consistent with
    ${\rco_1}$ and ${\rfr_1}$.  \ref{seq-le-rf-rf} follows from \ref{arm-rfi}.
    As noted \ref{lob-le}, $\rlob$ includes the order required by
    \ref{seq-kappa} and \ref{seq-le-delays-rf}.  
  \end{proof}
\end{theorem}

### 2021-05-26 paper/oopsla.tex
uncovered some examples
Added m5a

The proof of compilation to Arm in \jjr{} assumes that all internal reads can
be eliminated.
% Shortly after publication, \citet{anton}
% noticed a shortcoming of the implementation on \armeight{} in
% \jjr{\textsection 7}.  The proof given there assumes that all internal reads
% can be dropped.
However, this is not the case for acquiring reds.  For example, \jjr{}
disallows the following execution, where the final values of $x$ is $2$ and
the final value of $y$ is $2$.  This execution is allowed by \armeight{} and
\tso{}.
\begin{gather*}
  \PW{x}{2}\SEMI 
  \PR[\mACQ]{x}{r}\SEMI
  \PR{y}{s} \PAR
  \PW{y}{2}\SEMI
  \PW[\mREL]{x}{1}
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{2}}{}
      \raevent{b}{\DR[\mACQ]{x}{2}}{right=of a}
      \event{c}{\DR{y}{0}}{right=of b}
      \event{d}{\DW{y}{2}}{right=2.5em of c}
      \raevent{e}{\DW[\mREL]{x}{1}}{right=of d}
      \rf{a}{b}
      \sync{b}{c}
      \wk{c}{d}
      \sync{d}{e}
      \wk[out=-165,in=-15]{e}{a}
      % \rfi{a}{b}
      % \bob{b}{c}
      % \fre{c}{d}
      % \bob{d}{e}
      % \coe[out=-165,in=-15]{e}{a}
    \end{tikzinline}}
\end{gather*}
We discussed two approaches to this problem in \textsection\ref{sec:arm}.
% The solution we have adopted is to allow an acquiring read to be downgraded
% to a relaxed read when it is preceded (sequentially) by a relaxed write that
% could fulfill it.  This solution allows executions that are not allowed under
% \armeight{} since we do not insist that the local relaxed write is actually
% read from.  This may seem counterintuitive, but we don't see a local way to
% be more precise.

@@

@@

\section{Additional RMW Examples}
\begin{example}
  This definition ensures atomicity, disallowing executions such as
  \cite[Ex.~3.2]{DBLP:journals/pacmpl/PodkopaevLV19}:
  \begin{gather*}
    % \taglabel{RMW1}
    \begin{gathered}
      \PW{x}{0}\SEMI \PFADD[\mRLX][\mRLX]{x}{s}{1}
      \PAR
      \PW{x}{2}\SEMI \PR{x}{s}
      \\
      \hbox{\begin{tikzinline}[node distance=1.5em]
          \event{a2}{\DR{x}{0}}{}
          \event{a1}{\DW{x}{0}}{left=of a2}
          \rf{a1}{a2}
          \event{a3}{\DW{x}{2}}{right=of a2}
          \wk{a2}{a3}
          \event{b2}{\DW{x}{1}}{right=of a3}
          \event{b3}{\DR{x}{1}}{right=of b2}
          \rmw[out=-15,in=-165]{a2}[below]{b2}
          \wk{a3}{b2}
          \rf{b2}{b3}
          \liftrmw[out=165,in=15]{a3}{a2}
        \end{tikzinline}}
    \end{gathered}
  \end{gather*}
  By \ref{pom-rmw-atomic1}, since $\DWP{x}{2}\xwk\DWP{x}{1}$, it must be that
  $\DWP{x}{2}\xwk\DRP{x}{0}$, creating a cycle.
\end{example}

\begin{example}
  \label{ex:rmw-33}
  Two successful \RMW{}s cannot see the same write:
  \begin{gather*}
    \begin{gathered}
      \PW{x}{0}\SEMI (\FADD^{\mRLX,\mRLX}(\aLoc,1)\PAR \FADD^{\mRLX,\mRLX}(\aLoc,1))
      \\
      \hbox{\begin{tikzinline}[node distance=1.5em]
          \event{i}{\DW{x}{0}}{}
          \event{a1}{a{:}\DR{x}{0}}{right=3em of i}
          \event{a2}{b{:}\DW{x}{1}}{right=of a1}
          \event{b1}{c{:}\DR{x}{0}}{right=3em of a2}
          \event{b2}{d{:}\DW{x}{1}}{right=of b1}
          \rmw{a1}{a2}
          \rmw{b1}{b2}
          \rf{i}{a1}
          \rf[out=15,in=165]{i}{b1}
          \wk[out=-15,in=-165]{a1}{b2}
          \liftrmw[out=-15,in=-165]{a2}{b1}
          % \wk{a1}{b2}
          \wk{b1}{a2}
        \end{tikzinline}}
    \end{gathered}
  \end{gather*}
  The order from read-to-write is required by fulfillment.  
  Apply \ref{pom-rmw-atomic1} of the second \RMW{} to $a\xwk d$, we have that $a\xwk c$.  Subsequently
  applying \ref{pom-rmw-atomic2} of the first \RMW{}, we have $b \xwk c$, creating a cycle.
\end{example}

\begin{example}
  By using two actions rather than one, the definition allows examples such as the
  following, which is allowed by \armeight{} 
  \cite[Ex.~3.10]{DBLP:journals/pacmpl/PodkopaevLV19}:
  \begin{gather*}
    % \taglabel{RMW2}
    \begin{gathered}
      \PR{y}{r}\SEMI
      \PW{z}{r}
      \PAR
      \PR{z}{r}\SEMI
      \PW{x}{0}\SEMI
      \PFADD[\mRLX][\mREL]{x}{s}{1} \SEMI
      \PW{y}{s}{+}1
      \\[-1ex]
      \hbox{\begin{tikzinline}[node distance=1.5em]
          \event{a1}{\DR{y}{1}}{}
          \event{a2}{\DW{z}{1}}{right=of a1}
          \po{a1}{a2}
          \event{b1}{\DR{z}{1}}{right=3em of a2}
          \rf{a2}{b1}
          \event{b2}{\DW{x}{0}}{right=of b1}
          \event{b3}{\DR{x}{0}}{right=of b2}
          \rf{b2}{b3}
          \event{b4}{\DWRel{x}{1}}{right=2em of b3}
          \rmw{b3}{b4}
          \event{b5}{\DW{y}{1}}{right=of b4}
          \sync[out=-15,in=-165]{b1}{b4}
          \po[out=-20,in=-160]{b3}{b5}
          \rf[out=170,in=10]{b5}{a1}
        \end{tikzinline}}
    \end{gathered}
  \end{gather*}
\end{example}

\begin{example}
  Consider the \textsc{cdrf} example from \cite{DBLP:conf/pldi/LeeCPCHLV20}:
  \begin{gather*}
    \begin{gathered}
      \begin{aligned}
        &\PFADD[\mACQ][\mREL]{x}{r}{1}\SEMI \IF{r{=}0}\THEN \PW{y}{1} \FI
        \\\PAR\;\;&
        \PFADD[\mACQ][\mREL]{x}{r}{1}\SEMI \IF{r{=}0}\THEN \IF{y}\THEN \PW{x}{0} \FI \FI
      \end{aligned}
      \\
      \hbox{\footnotesize\begin{tikzinline}[node distance=1.5em]
          \raevent{a1}{\DR[\mACQ]{x}{0}}{}
          \raevent{a1b}{\DW[\mREL]{x}{1}}{right=of a1}
          \event{a2}{\DW{y}{1}}{right=of a1b}
          \raevent{b0}{\DR[\mACQ]{x}{0}}{right=3em of a2}
          \raevent{b0b}{\DW[\mREL]{x}{1}}{right=of b0}
          \event{b1}{\DR{y}{1}}{right=of b0b}
          \event{b2}{\DW{x}{0}}{right=of b1}
          \rmw{a1}{a1b}
          \rmw{b0}{b0b}
          \rf[out=-13,in=-163]{a2}{b1}
          \sync[out=20,in=160]{a1}{a2}
          \sync[out=20,in=160]{b0}{b1}
          \po{b1}{b2}
          \rf[out=-165,in=-12]{b2}{a1}
        \end{tikzinline}}
    \end{gathered}
  \end{gather*}
\end{example}
\begin{example}
  Consider this example from \cite[\textsection C]{DBLP:conf/pldi/LeeCPCHLV20}:
  \begin{gather*}
    \begin{gathered}
      \begin{aligned}
        &\PCAS[\mRLX][\mRLX]{x}{r}{0}{1}\SEMI \IF{r{\leq}1}\THEN \PW{y}{1} \FI
        \\\PAR\;\;&
        \PCAS[\mRLX][\mRLX]{x}{r}{0}{2}\SEMI \IF{r{=}0}\THEN \IF{y}\THEN \PW{x}{0} \FI \FI
      \end{aligned}
      \\
      \hbox{\footnotesize\begin{tikzinline}[node distance=1.5em]
          \event{a1}{\DR{x}{0}}{}
          \event{a1b}{\DW{x}{1}}{right=of a1}
          \event{a2}{\DW{y}{1}}{right=of a1b}
          \event{b0}{\DR{x}{0}}{right=3em of a2}
          \event{b0b}{\DW{x}{2}}{right=of b0}
          \event{b1}{\DR{y}{1}}{right=of b0b}
          \event{b2}{\DW{x}{0}}{right=of b1}
          \rmw{a1}{a1b}
          \rmw{b0}{b0b}
          \rf[out=-13,in=-163]{a2}{b1}
          \po[out=20,in=160]{a1}{a2}
          \po[out=20,in=160]{b0}{b1}
          \po{b1}{b2}
          \rf[out=-165,in=-12]{b2}{a1}
        \end{tikzinline}}
    \end{gathered}
  \end{gather*}
\end{example}

@@

\myparagraph{Stupid Bug in LDRF}
Definition of race is wrong.
Should say that at least one is relaxed.

>>>>

The proof of compilation to Arm in \jjr{} assumes that all internal reads can
be eliminated.
% Shortly after publication, \citet{anton}
% noticed a shortcoming of the implementation on \armeight{} in
% \jjr{\textsection 7}.  The proof given there assumes that all internal reads
% can be dropped.
However, this is not the case for acquiring reads.  For example, \jjr{}
disallows the following execution, where the final values of $x$ is $2$ and
the final value of $y$ is $2$.  This execution is allowed by \armeight{} and
\tso{}.
\begin{gather*}
  \PW{x}{2}\SEMI 
  \PR[\mACQ]{x}{r}\SEMI
  \PR{y}{s} \PAR
  \PW{y}{2}\SEMI
  \PW[\mREL]{x}{1}
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{2}}{}
      \raevent{b}{\DR[\mACQ]{x}{2}}{right=of a}
      \event{c}{\DR{y}{0}}{right=of b}
      \event{d}{\DW{y}{2}}{right=2.5em of c}
      \raevent{e}{\DW[\mREL]{x}{1}}{right=of d}
      \rf{a}{b}
      \sync{b}{c}
      \wk{c}{d}
      \sync{d}{e}
      \wk[out=-165,in=-15]{e}{a}
      % \rfi{a}{b}
      % \bob{b}{c}
      % \fre{c}{d}
      % \bob{d}{e}
      % \coe[out=-165,in=-15]{e}{a}
    \end{tikzinline}}
\end{gather*}
We discussed two approaches to this problem in \textsection\ref{sec:arm}.
% The solution we have adopted is to allow an acquiring read to be downgraded
% to a relaxed read when it is preceded (sequentially) by a relaxed write that
% could fulfill it.  This solution allows executions that are not allowed under
% \armeight{} since we do not insist that the local relaxed write is actually
% read from.  This may seem counterintuitive, but we don't see a local way to
% be more precise.

@@

\myparagraph{Data Race Freedom}

The definition of data race is wrong in \jjr{}.  It should require that that
at least one action is relaxed.

Not that the definition of \emph{$L$-stable} applies in the case that
conflicting writes are totally ordered.  This gives a result more in the
spirit of \cite{Dolan:2018:BDR:3192366.3192421}.  In particular, this special
case of the theorem clarifies the discussion of \ref{past}.

@@

@@
